Doing Better Than UCT: \\ Rational Monte Carlo Sampling in Trees

### 2011-08-13 paper/rmcts/paper.tex
algorithm fixed

Such strategies are used to select the best arm, and, by extension,
the best action in Monte Carlo tree search. Monte Carlo tree search is
used to solve Markov Decision Processes (MDP) approximately. An MDP is
defined by the set of states $S$, the set of actions $A$, the
transition table $T(s, a, s')$, the reward table $R(s, a, s')$, the
initial state $s$ and the goal state $t$: $(S, A, T, R, s, t)$.  Monte
Carlo tree search explores an MDP by \emph{rollouts}---trajectories
from the current state to a state in which a termination condition is
satisfied (either the goal state, or a cutoff state for which the
reward is evaluated approximately). Algorithm UCB that maximizes the
accumulated reward \cite{Auer.ucb} had been extended to tree search as
UCT \cite{Kocsis.uct}.

@@

\begin{algorithm}[t]
\caption{Two-stage Monte-Carlo tree search sampling}
\label{alg:two-stage-mcts}
\begin{algorithmic}[1]
\Procedure{Rollout}{node}
  \State $depth \gets 0$
  \State $reward \gets 0$
  \While {$\lnot leaf(node)$}
    \If {$depth=1$}
      \State $action \gets$\Call{FirstAction}{node} \Comment minimizes simple regret, e.g. $\frac 1 2$-greedy
    \Else
      \State $action \gets$\Call{NextAction}{node} \Comment  UCB --- minimizes accumulated regret
    \EndIf
    \State $node' \gets$\Call{NextState}{node, action}
    \State $reward \gets reward+$\Call{TransitionReward}{node, action, node'}
    \State $node \gets node'$
    \State $depth \gets depth+1$
  \EndWhile
  \State \textbf{return} $reward$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

The improvement is based on the notion of value of information (VOI),
but VOI is used implicitly in the analysis of the algorithm, rather
than computed or learned explicitly in order to plan the rollouts. A
further improvement can be achieved by computing or estimating the VOI
of the rollouts and chosing a rollout that maximizes the VOI. Using VOI
to guide Monte Carlo sampling instead of relying on the sample means
of different actions  is particularly beneficial when different
actions have qualitatively different distributions: for example, when the
distribution variance varies signficantly between the actions.

@@

That is, the case $\frac {\partial P_j} {\partial \varepsilon}=0$
corresponds to \emph{0.5-greedy sampling}, and for a large number of
arms $K$ 
\begin{equation}
\mathbb{B}r_{1/2\mbox{-}greedy}\approx \mathbb{B}r_{uniform}^2
\label{eq:derive-uniform-vs-half-greedy}
\end{equation}
Equation (\ref{eq:derive-uniform-vs-half-greedy}) suggests that
0.5-greedy sampling has a much higher convergence rate than uniform
sampling for a large number of arms.

>>>>

Such strategies are used to select the best arm, and, by extension,
the best action in Monte Carlo tree search. Monte Carlo tree search is
used to solve Markov Decision Processes (MDP) approximately. An MDP is
defined by the set of states $S$, the set of actions $A$, the
transition table $T(s, a, s')$, the reward table $R(s, a, s')$, the
initial state $s$ and the goal state $t$: $(S, A, T, R, s, t)$
\cite{Russell.aima}.  Monte Carlo tree search explores an MDP by
\emph{rollouts}---trajectories from the current state to a state in
which a termination condition is satisfied (either the goal state, or
a cutoff state for which the reward is evaluated
approximately). Algorithm UCB that maximizes the accumulated reward
\cite{Auer.ucb} had been extended to tree search as UCT
\cite{Kocsis.uct}.

@@

\begin{algorithm}[t]
\caption{Two-stage Monte-Carlo tree search sampling}
\label{alg:two-stage-mcts}
\begin{algorithmic}[1]
\Procedure{Rollout}{node, depth=1}
  \If {\Call{IsLeaf}{node, depth}}
    \State \textbf{return} 0
  \Else
    \If {depth=1}
      \State action $\gets$ \Call{FirstAction}{node} \Comment minimizes simple regret, e.g. $\frac 1 2$-greedy
    \Else
      \State action $\gets$ \Call{NextAction}{node} \Comment  UCB --- minimizes accumulated regret
    \EndIf
    \State next-node $\gets$ \Call{NextState}{node, action}
    \State reward $\gets$ \Call{TransitionReward}{node, action, next-node}
    \State \hspace{4em} + \Call{Rollout}{next-node, depth+1} \Comment call {\sc Rollout} recursively
    \State \Call{UpdateStats}{node, action, reward}
  \EndIf
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

The improvement is inspired by the notion of value of information (VOI),
but VOI is used implicitly in the analysis of the algorithm, rather
than computed or learned explicitly in order to plan the rollouts. A
further improvement can be achieved by computing or estimating the VOI
of the rollouts and chosing a rollout that maximizes the VOI. Using VOI
to guide Monte Carlo sampling instead of relying on the sample means
of different actions  is particularly beneficial when different
actions have qualitatively different distributions: for example, when the
distribution variance varies signficantly between the actions.

@@

That is, the case $\frac {\partial P_j} {\partial \varepsilon}=0$
corresponds to \emph{$\frac 1 2$-greedy sampling}, and for a large number of
arms $K$ 
\begin{equation}
\mathbb{B}r_{1/2\mbox{-}greedy}\approx \mathbb{B}r_{uniform}^2
\label{eq:derive-uniform-vs-half-greedy}
\end{equation}
Equation (\ref{eq:derive-uniform-vs-half-greedy}) suggests that
$\frac 1 2$-greedy sampling has a much higher convergence rate than uniform
sampling for a large number of arms.
